A tool set for random number generation on GPUs in R

### 2021-11-16 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
update paper1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2972 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.154   0.003   0.588
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   7.388   0.432   7.814
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

month_gpu<-vclMatrix(month,type="integer")
system.time(result_month <- fisher.sim(month_gpu, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(256,64)))
##    user  system elapsed 
##   1.047   0.016   2.597
result_month$simNum
## [1] 1015808
result_month$counts
## [1] 410825
result_month$p.value
## [1] 0.4044323
\end{verbatim}
\end{kframe}
\end{knitrout}
We got 409524 cases whose test statistics are below the observed threshold when there are an actual number of 1015808 simulations executed, so the p-value from \fct{clrng::fisher.sim} for this table is about 0.40443, which is close enough to the p-value from \fct{stats::fisher.test}. \fct{clrng::fisher.sim} takes about 0.69 seconds, \pkg{clrng} accounts for 4.64\% of the elapsed time on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1015808)$p.value
## [1] 0.4036733
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1015808))
##    user  system elapsed 
##  18.468   0.016  18.468
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
observed_w= -clrng::logfactSum(week, c(16,16))/almost.1
week_GPU<-gpuR::vclMatrix(week,type="integer")
result_weekgpu<-clrng::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,64))
result_weekgpu$simNum
## [1] 10010624
result_weekgpu$counts
## [1] 1193
result_weekgpu$p.value
## [1] 0.0001192733
\end{verbatim}
\end{kframe}
\end{knitrout}

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
system.time(clrng::fisher.sim(week_GPU, 1e7, streams=streams,
            type="double", returnStatistics=FALSE,Nglobal = c(256,64)))
##    user  system elapsed 
##   0.887   0.020  10.614
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
fisher.test(week,simulate.p.value = TRUE,B=10010624)$p.value
## [1] 0.0001249672
system.time(fisher.test(week,simulate.p.value = TRUE,B=10010624))
##    user  system elapsed 
##  64.151   0.080  64.172
\end{verbatim}
\end{kframe}
\end{knitrout}

>>>>

We generate a large-size matrix of 100 million double-precision Gaussian random numbers, and compare the run-time between using \fct{stats::rnorm} and \fct{clrng::rnorm}. The best performance we have seen using \fct{clrng::rnorm} is around 120 times faster with $512 \times 128$ work-items than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- createStreamsGpu(n = 512 * 128)
system.time(clrng::rnorm(c(10000,10000), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.155   0.000   0.585
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(matrix(stats::rnorm(10000^2),10000,10000))
##    user  system elapsed 
##   7.388   0.432   7.814
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

month_gpu<-vclMatrix(month,type="integer")
system.time(result_month <- fisher.sim(month_gpu, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = c(256,64)))
##    user  system elapsed 
##   1.057   0.004   2.596
result_month$simNum
## [1] 1015808
result_month$counts
## [1] 410825
result_month$p.value
## [1] 0.4044323
\end{verbatim}
\end{kframe}
\end{knitrout}
We got 409524 cases whose test statistics are below the observed threshold based on an actual number of 1015808 simulations executed, so the p-value from \fct{clrng::fisher.sim} for this table is about 0.40443, which is close enough to the p-value from \fct{stats::fisher.test}. \fct{clrng::fisher.sim} takes about 0.69 seconds, \pkg{clrng} accounts for 4.64\% of the elapsed time on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
system.time(result_monthcpu<-stats::fisher.test(month,simulate.p.value = TRUE,B=1015808))
##    user  system elapsed 
##  18.606   0.027  18.626
result_monthcpu$p.value
## [1] 0.4036733
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
observed_w= -clrng::logfactSum(week, c(16,16))/almost.1
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(result_weekgpu<-clrng::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = c(256,64)))
##    user  system elapsed 
##   0.925   0.012  10.653
result_weekgpu$simNum
## [1] 10010624
result_weekgpu$counts
## [1] 1193
result_weekgpu$p.value
## [1] 0.0001192733
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.0001249672, which should require a larger number of simulations to get a more accurate p-value. With more than ten million simulations, we get 1205 cases and a p-value around 0.000120472 by \fct{fisher.sim}. \fct{stats::fisher.test} takes 91.6 seconds, while \fct{fisher.sim} takes about 4.282 seconds, the elapsed time is decreased to about 4.67\% of the time taken on CPU.

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
system.time(result_weekcpu<-fisher.test(week,simulate.p.value = TRUE,B=10010624))
##    user  system elapsed 
## 117.213   0.152 117.289
result_weekcpu$p.value
## [1] 0.0001249672
\end{verbatim}
\end{kframe}
\end{knitrout}
